A weakly structured stem for human origins in Africa

### 2022-02-24 paper.tex
cleanup upon read-through

Regardless of the model choice for early epochs, inference of human demographic
history for the last 150kya was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135kya, with low to moderate levels of
subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140kya. 
We conclude that geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (see below). Fixed parameters included
the time of divergence between Western and Eastern African populations,
set to 60kya, just prior to the split of Eurasians and East Africans set to 50kya. 
We also fixed the amount of admixture from Neanderthals to Europeans directly
following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). Such constraints, informed by prior research,
allowed us to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
($m\approx2\times10^{-4}$). We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE} (Figure~\ref{fig:1}).

@@

To account for population structure prior to 135kya, three of our four models allowed for two or more
“stem” populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase. 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or “merger”) events from the
other stem population, so that recent populations are formed by mergers of multiple ancestral populations. Depending on parameter values,
this scenario encompasses archaic introgression and
fragmentation-and-coalescence models. For many parameters, confidence intervals based on bootstrapping are relatively narrow
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. Model assumptions
have, however, a larger impact on parameter estimates (and thus, uncertainty). To convey model uncertainty, we
highlight features of the two inferred models with high likelihoods. These are referred to as the
“multiple-merger” and the the “continuous-migration” model. Both allow for
migration between stem branches, but differ primarily in the timing of the
early divergence of stem populations and their relative $N_e$
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration ($LL \approx -102,600$ vs.
$-107,700$ in the merger model and $LL \approx -115,500$ vs. $-126,600$ with
continuous migration). With continuous migration between stems, population
structure extends back to 1.1--1.4Mya (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is similar
to inferred migration rates between connected contemporary populations over the
past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic admixture models (Figure~\ref{fig:supp-possible-models}C) as
the early branches remain closely related (Figure 3). Because of this 
relatedness, only $1\%$ to $4\%$ of genetic differentiation among modern populations can be traced back
to this early population structure (Figure 3 and supplementary text)

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into “Stem 1E” and “Stem 1S”
479kya. The timing of this divergence was also poorly constrained ($\sigma_T=
166$kya). These populations evolve independently until approximately 119kya
when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the Western Africans and
Eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25kya. The later Stem
2 contribution to the Western African Mende resulted in better model fits (LL).
This may indicate that an ancestral Stem 2 population occupied Western or
Central Africa, broadly speaking. The differing proportions in the Nama and
Eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

Previous studies have found support for archaic admixture in Africa using
two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
(cSFS) \citep{Durvasula2020-td}, and reconstruction of gene genealogies
\citep{Speidel2019-nj}. However, none of these studies considered a weakly
structured stem. We validated our inferred models with additional independent
approaches. We find that the observed cSFS (conditioned on the derived allele
being carried in the Neanderthal sample) is very well-described by the merger
model (Figures~\ref{fig:4}A-C and
\ref{fig:supp-csfs-single-origin}--\ref{fig:supp-csfs-merger-with-stem-migration}),
even though this statistic was not used in the fit. Both models outperform
archaic models fit directly to the cSFS (for example, compare with Figure~1 in
\citet{Durvasula2020-td}).

@@

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time, which are commonly interpreted as effective
population size histories. Trajectories of $N_e$ from prior studies strongly
support a bottleneck in Out of Africa populations. African populations also dip
in effective size around 100-200kya, although this decrease is more modest than
the OOA event \citep{Mallick2016-lx}. One possible explanation is that
ancestral population structure during the Middle Stone Age inflated $N_e$ and
subsequent gene flow among ancestral Africans resulted in a more recent
decreased $N_e$ \citep{Mazet2016-wn}.
%To our knowledge, explicit comparisons of this hypothesis with SMC-style
%approaches have not been published.
We simulated genomic data under four inferred demographic models to compare
with observed coalescence rate histories. All models, including the
single-origin model, generate multi-modal curves indicating that the ancestral
increase in $N_e$ between 100kya-1Mya is a general feature of the coalescent
rate in humans, although none of the models perfectly match the exact timing of
the early fluctuations in $N_e$ observed in the data.
%Simulated Relate Ne curves under the CM and MM are distinguished by having 2
%and 3 modes, respectively.  The Continuous Migration model had the better
%qualitative fit with the data, but the ancestral mode was shifted further back
%in time relative to the data.
Each simulated model also underestimated the level of population growth during
the Holocene, indicating additional free parameters are necessary to refine
these growth rates.

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single origin model must be explained by archaic
admixture. This is also true of this study. While it remains prohibitive to
fully explore the space of plausible models of early human population structure,
we sought to capture model uncertainty by exploring multiple parameterizations of
early history.
\bmhcomment{we use this phrase a lot. human or hominin? Tim?}
The best-fit models presented here include reticulation and migration between
early human populations rather than archaic admixture from long-isolated branches.
We cannot rule out that more complex models involving
additional stems, or hybrid models including both weak structure and archaic
admixture may better explain the data. Because parameters related to the split
time, migration rates, and relative sizes of the early stems were variable
across models, reflecting a degree of confounding among these parameters, we
refrained from introducing additional branches associated with more
parameters during that period.
\sgcomment{Some statement about hitting the boundary}
Rather than interpreting the two stems as representing well-defined
and stable populations over hundreds of thousands of years, we interpret the
weakly structured stem as consistent with a population coalescence and
fragmentation model \citep{Scerri2019-xg}.
Models including additional diversity within Africa,
and ancient DNA samples from Africa, could further distinguish
the archaic admixture model from the weakly-structured-stem model.

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S from southern Africa moved northward thereby encountering the Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
modern groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population ~120kya.
The second ~100kya between Stem 1E and Stem 2, results in the formation of the
ancestors of East/West Africans as well as later “Out of Africans”. The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested for the
Paleo-Agulhas plain \citep{Marean2014-pg}.

Following these merger events, the stems subsequently fracture into
subpopulations which then appear to persist over the past ~120kya. These
subpopulations can be linked to modern-day groups despite subsequent gene flow
across the continent; for example, a genetic lineage sampled in the Nama has an
XX probability of being in the ancestral ‘southern’ subpopulation 50kya versus
XX probability of being in the ‘eastern’ subpopulation (Figure XX). We also
find that Stem 2 continued to contribute to western Africans during the Last
Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table 1).

@@

Multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}. If we
assume that this correspondence also holds for early human ancestral
populations in Africa, the weakly-structured-stem model would predict that the
fossils deriving from each of the two stems should be morphologically quite
similar (supplementary text X)\aprcomment{Will we still include this?}, even if
they are from different parts of Africa, with differences across stems closer
to those across contemporary human populations than to human-Neanderthal
differences. Differences within stems would be comparable to differences within
recent human populations. Identifying the morphological signature of ancient
structure in the fossil record will require careful analysis of many broadly
similar remains. Our model further predicts that morphologically distinct
crania such as XX and YY are unlikely representatives of branches that
contributed appreciably to modern human ancestries.

>>>>

Regardless of the model choice for early epochs, inference of human demographic
history for the last 150kya was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135kya, with low to moderate levels of
subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140kya. 
We conclude that geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto the more ancient
`stem' groups as only a small proportion of drift between modern populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events. 
Fixed parameters included
the time of divergence between Western and Eastern African populations,
set to 60kya, just prior to the split of Eurasians and East Africans set to 50kya. 
We also fixed the amount of admixture from Neanderthals to Europeans directly
following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). These constraints allowed us to integrate information from previous \sgcomment{genetic and archaelogical?} research,
to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
($m\approx2\times10^{-4},$ the proportion of migrant lineages per generation). We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE} (Figure~\ref{fig:1}).

@@

To account for population structure prior to 135kya, three of our four models allowed for two or more
“stem” populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase. 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or “merger”) events from the
other stem population, so that recent populations are formed by mergers of multiple ancestral populations. Depending on parameter values,
this scenario encompasses archaic introgression and
fragmentation-and-coalescence models. For many parameters, confidence intervals based on bootstrapping are relatively narrow
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. Model assumptions
have, however, a larger impact on parameter estimates (and thus, real uncertainty). To convey model uncertainty, we
highlight features of the two inferred models with high likelihoods. These are referred to as the
“multiple-merger” and the the “continuous-migration” models. Both allow for
migration between stem branches, but differ primarily in the timing of the
early divergence of stem populations and their relative $N_e$
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration ($LL \approx -102,600$ vs.
$-107,700$ in the merger model and $LL \approx -115,500$ vs. $-126,600$ with
continuous migration). With continuous migration between stems, population
structure extends back to 1.1--1.4Mya (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is similar
to inferred migration rates between connected contemporary populations over the
past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic admixture models (Figure~\ref{fig:supp-possible-models}C) as
the early branches remain closely related, \sgcomment{and each branch contributes large amounts to all contemporary populations} (Figure 3). 
Because of this 
relatedness, only $1\%$ to $4\%$ of genetic differentiation among modern populations can be traced back
to this early population structure (Figure 3 and supplementary text)

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into “Stem 1E” and “Stem 1S”
479kya. The timing of this divergence was also poorly constrained ($\sigma_T=
166$kya). These populations evolve independently until approximately 119kya
when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the Western Africans and
Eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25kya. The later Stem
2 contribution to the Western African Mende resulted in better model fits (LL \sgcomment{Something is missing here}).
This may indicate that an ancestral Stem 2 population occupied Western or
Central Africa, broadly speaking. The differing proportions in the Nama and
Eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

Previous studies have found support for archaic admixture in Africa using
two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
(cSFS) \citep{Durvasula2020-td}, and reconstruction of gene genealogies
\citep{Speidel2019-nj}. However, none of these studies considered a weakly
structured stem. We validated our inferred models with additional independent
approaches. We find that the observed cSFS (conditioned on the derived allele
being carried in the Neanderthal sample) is very well-described by the merger
model (Figures~\ref{fig:4}A-C and
\ref{fig:supp-csfs-single-origin}--\ref{fig:supp-csfs-merger-with-stem-migration}),
even though this statistic was not used in the fit. Both models actually outperform
archaic models fit directly to the cSFS (for example, compare with Figure~1 in
\citet{Durvasula2020-td}).

@@

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time, which are commonly interpreted as effective
population size histories. Trajectories of $N_e$ from prior studies strongly
support a bottleneck in Out of Africa populations. African populations also dip
in effective size around 100-200kya, although this decrease is more modest than
the OOA event \citep{Mallick2016-lx}. One possible explanation is that
ancestral population structure during the Middle Stone Age inflated $N_e$ and
subsequent gene flow among ancestral Africans resulted in a more recent
decreased $N_e$ \citep{Mazet2016-wn}.
%To our knowledge, explicit comparisons of this hypothesis with SMC-style
%approaches have not been published.
We simulated genomic data under four inferred demographic models to compare
with observed coalescence rate histories. All models, including the
single-origin model, generate multi-modal curves indicating that the ancestral
increase in $N_e$ between 100kya-1Mya is a general feature of the coalescent
rate in humans, although none of the models perfectly match the exact timing of
the early fluctuations in $N_e$ observed in the data.
%Simulated Relate Ne curves under the CM and MM are distinguished by having 2
%and 3 modes, respectively.  The Continuous Migration model had the better
%qualitative fit with the data, but the ancestral mode was shifted further back
%in time relative to the data.
Each simulated model also underestimated the level of population growth during
the Holocene, indicating additional free parameters are necessary to refine
these growth rates. \sgcomment{I am not sure I am happy with these interpretations 
-- let's discuss. I feel like this paragraph does not support our data well, and a likely culprit 
in explaining the differences is the inaccuracy of TMRCA estimates from real data.}

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single origin model must be explained by archaic
admixture. This is also true of this study. While it remains prohibitive to
fully explore the space of plausible models of early human population structure,
we sought to capture model uncertainty by exploring multiple parameterizations of
early history.
\bmhcomment{we use this phrase a lot. human or hominin? Tim?}
The best-fit models presented here include reticulation and migration between
early human populations rather than archaic admixture from long-isolated branches.
We cannot rule out that more complex models involving
additional stems, or hybrid models including both weak structure and archaic
admixture may better explain the data. Because parameters related to the split
time, migration rates, and relative sizes of the early stems were variable
across models, reflecting a degree of confounding among these parameters, we
refrained from introducing additional branches associated with more
parameters during that period.
Rather than interpreting the two stems as representing well-defined
and stable populations over hundreds of thousands of years, we interpret the
weakly structured stem as consistent with a population coalescence and
fragmentation model \citep{Scerri2019-xg}.
Models including additional diversity within Africa,
and early ancient DNA samples from Africa, could further distinguish
the archaic admixture model from the weakly-structured-stem model.

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S from southern Africa moved northward thereby encountering the Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
modern groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population ~120kya.
The second $\approx100$kya between Stem 1E and Stem 2, results in the formation of the
ancestors of East/West Africans as well as later “Out of Africans”. The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested for the
Paleo-Agulhas plain \citep{Marean2014-pg}.

Following these merger events, the stems subsequently fracture into
subpopulations which then appear to persist over the past ~120kya. These
subpopulations can be linked to modern-day groups despite subsequent gene flow
across the continent; for example, a genetic lineage sampled in the Nama has an
\sgcomment{XX} probability of being inherited from the ancestral ‘southern’ subpopulation 50kya versus
\sgcomment{XX} probability of being inherited from the ‘eastern’ subpopulation (Figure \sgcomment{XX}). We also
find that Stem 2 continued to contribute to western Africans during the Last
Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table 1).

@@

Multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}. If we
assume that this correspondence also holds for early human ancestral
populations in Africa, the weakly-structured-stem model would predict that the
fossils deriving from each of the two stems should be morphologically quite
similar (supplementary text \sgcomment{X})\aprcomment{Will we still include this?} \sgcomment{yes!}, even if
they are from different parts of Africa, with differences across stems closer
to those across contemporary human populations than to human-Neanderthal
differences. Identifying the morphological signature of ancient
structure in the fossil record will thus require careful analysis of many broadly
similar remains. Morphologically distinct
crania such as \sgcomment{XX} and \sgcomment{YY} are unlikely representatives of branches that
contributed appreciably to modern human ancestries.
